Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-02 doc/papers/cgo2014/paper.tex
Add bench part

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. The Listing~\ref{lst:rosen} 
illustrates the use of a small set of functions: a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed with operator fusion.

@@

Fortran/C/C++ have been at the hearth of high performance computing for many years.
Compiler infrastructures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
% You are speaking about user written assembly code? Or generated code?
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
good abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of third-party libraries like GotoBlas~\cite{gotoblas2008}.
% pas de nt2?

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function as side effects on its argument or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially easy in Python thanks to the
ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

A first approach would be to use recursive tasks and manual loop chunking.
\todo{pierrick ? c'est bête comme idée ?}
% Les taches sont en général plus lente que les boucles for (malheureusement)
% et il y a des fois ou on ne pourra jamais être plus rapide si on passe par
% des taches.
% Souvent, c'est a cause du scheduling static qui permet d'optimiser le cache
% et aussi parce que les compilos, quand ils voient un omp for static, ils
% font directement du déroulage de boucle. (Jusqu'au full unrolling des fois
% avec autant de tache que de thread). 
%
% Dans notre cas, on ne connait pas forcement la taille de l'espace d'iteration
% a la compilation. On ne pourra donc pas faire trop de unrolling mais bon, on
% doit pouvoir faire comme gcc. Par contre, pour l'optimisation du cache, les
% boucles for assurent que c'est le même noeud qui exécute le même "chunk" entre
% les boucles for et ça, on le perd avec les taches. Je ne sais pas si ça joue
% vraiment.
% Autre problem. On va créé des taches en assez grande quantité. Ce que libgomp
% supporte assez mal. Aussi, on va créé du datasharing pour des petites taches
% (iteration de longueur 1 dans le cas du (1, 1000000)) ce qui est couteux et
% assez mal supporté par libgomp. Il faudra voir comment ça se passe avec libiomp
% qui est utilisé par clang.
%
% Au final : Ca doit être jouable. Il y a plusieurs raisons qui peuvent faire
% qu'on a de mauvaise perfs mais sans mesures, on ne sait pas.

@@

% The first one holds 8 AMD Magny Cours processors for a
% total of 48 cores. Each core has access to 64\,KB of L1 cache,
% 512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
% is shared between the 6 cores of a processor. This configuration
% provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
% up to SSE4. We
% will refer to this configuration as \textbf{AMD48}.
% On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
% from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
% from OpenMP improvement. This clang version is based on clang 3.4 release
% (LLVM:233b1e3f034, clang-omp:6f358bcd87).

% The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a ttal of
% 8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
% 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
% is shared between the 8 processors. This configuration
% provides a total of 63\,GB of main memory. It supports up to AVX. We will
% refer to this configuration as \textbf{Intel8}.
% On Intel8, backend compilers are gcc 4.9 (20140528) and the clang-omp version
% from intel.

@@

The ideas described in this paper are implemnted in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The architecture of the compiler is briefely described
in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
the Pythran subset. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimization then outputs either
Python code, in a source-to-source fashion, or C++ template code that makes
call to the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instanciate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

@@

@@

\subsection{Results and Analysis}

>>>>

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This is a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. The Listing~\ref{lst:rosen}
illustrates the use of a small set of functions: a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed with operator fusion.

@@

Fortran/C/C++ have been at the hearth of high performance computing for many years.
Compiler infrastructures for these languages also have a long history and
implement many optimizations on the control flow graph, basic blocks or
loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
good abstractions for explicit vectorization. Many highly tuned kernels are
available in the form of third-party libraries like GotoBlas~\cite{gotoblas2008}.
% pas de nt2?

@@

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its argument or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially easy in Python thanks to the
ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

%A first approach would be to use recursive tasks and manual loop chunking.
%\todo{pierrick ? c'est bête comme idée ?}
% Les taches sont en général plus lente que les boucles for (malheureusement)
% et il y a des fois ou on ne pourra jamais être plus rapide si on passe par
% des taches.
% Souvent, c'est a cause du scheduling static qui permet d'optimiser le cache
% et aussi parce que les compilos, quand ils voient un omp for static, ils
% font directement du déroulage de boucle. (Jusqu'au full unrolling des fois
% avec autant de tache que de thread).
%
% Dans notre cas, on ne connait pas forcement la taille de l'espace d'iteration
% a la compilation. On ne pourra donc pas faire trop de unrolling mais bon, on
% doit pouvoir faire comme gcc. Par contre, pour l'optimisation du cache, les
% boucles for assurent que c'est le même noeud qui exécute le même "chunk" entre
% les boucles for et ça, on le perd avec les taches. Je ne sais pas si ça joue
% vraiment.
% Autre problem. On va créé des taches en assez grande quantité. Ce que libgomp
% supporte assez mal. Aussi, on va créé du datasharing pour des petites taches
% (iteration de longueur 1 dans le cas du (1, 1000000)) ce qui est couteux et
% assez mal supporté par libgomp. Il faudra voir comment ça se passe avec libiomp
% qui est utilisé par clang.
%
% Au final : Ca doit être jouable. Il y a plusieurs raisons qui peuvent faire
% qu'on a de mauvaise perfs mais sans mesures, on ne sait pas.

@@

@@

The ideas described in this paper are implemented in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
the Pythran subset. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimization then outputs either
Python code, in a source-to-source fashion, or C++ template code that makes
call to the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instantiate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

@@

As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate these optimisation. It is use with Pythran (commit: 488b7ba)
with optimization enabled/disabled. Finally, we compare result with Pypy (???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

Benchmarking is done one two architecture with different backend compilers.

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We will refer to this configuration as \textbf{AMD48}.
On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
from OpenMP improvement. This clang version is based on clang 3.4 release
(LLVM:233b1e3f034, clang-omp:6f358bcd87).

The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 processors. This configuration
provides a total of 63\,GB of main memory. It supports up to AVX. We will
refer to this configuration as \textbf{Intel8}.
On Intel8, backend compilers is gcc 4.9 (20140528).

@@

% For performance comparison between different Pythran version
To compare optimisation result, we run the numpy-benchmarks test suite multiple
time on the rosen.py benchmark. Results on Figure~\ref{rosenamd48.png} and \ref{rosenintel8.png} show difference
one speed which could be explain by AVX support on the intel8 architecture while amd48
support up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{rosen_amd48.png}
\caption{Rosen on AMD48}
\label{fig:rosenamd48}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{rosen_intel8.png}
\caption{Rosen on Intel8}
\label{fig:rosenintel8}
\end{figure}

\subsection{Comparison with others Python compiler}

Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
use of technique described in this paper. As we can see, both SIMD instruction
and parallelization have an impact on execution time. We can notice that SIMD
contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
Also, we can notice that gcc as a backend compiler have more optimization opportunity.

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{arc_distance_amd48.png}
\caption{Arc distance on AMD48}
\label{fig:arc_distance}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{arc_distance_intel8.png}
\caption{Arc distance on Intel8}
\label{fig:arc_distance_intel8}
\end{figure}

At the opposite, Figure~\ref{fig:diffusion} show that clang is better when generated
code make a more intensive use of templates (heavily use with gsliced\_ndarray).
gsliced\_ndarray is a non aligned view on an array in this case so SIMD optimization
can't be applied on the second dimension even if loop fusion and temporary
elimination is performed.  We can notice that Parakeet perform really well on
extensive gslice use.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{diffusion_amd48.png}
\caption{Diffusion on AMD48}
\label{fig:diffusion}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{diffusion_intel8.png}
\caption{Diffusion on Intel8}
\label{fig:diffusion_intel8}
\end{figure}

Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
is correct for these kind of functions.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{l2norm_amd48.png}
\caption{L2norm on AMD48}
\label{fig:l2norme}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{l2norm_intel8.png}
\caption{L2norm on Intel8}
\label{fig:l2norm_intel8}
\end{figure}
